New insights into the Weddell Sea ecosystem applying a network approach

### 2022-12-19 Manuscript/Manuscript.Rmd
LAS - Reviewed up to results

<!-- Intro & advantages of studying food webs -->
Building a network of feeding interactions among species is a fundamental step to understand ecological communities and is a valuable tool to manage ecosystems for conservation of biodiversity and functioning [@Thompson2012]. Although analyses considering presence/absence data on predator-prey interactions have provided insights about the structure and functioning of communities [e.g. @Kortsch2015; @Marina2018; @Cordone2020; @Rodriguez2022], more information is needed to effectively charaterise the ecosystems' dynamics and stability. In this regard, the intensity of the interactions between species or the interaction strength became of paramount importance in recent years [@Carrara2015; @Allesina2015; @Nilsson2016].

<!-- Interaction strength -->
Several methodologies have been applyied to estimate the interaction strength in food webs, where the quantity and quality of data mostly determines which approach is more convenient [@Berlow2004].

<!-- General description of the Weddell Sea -->
The Weddell Sea represents the southerly part of the Atlantic Sector of the Southern Ocean. About one quarter of the Weddell Sea’s entire marine area lies over the continental shelf, stretching along the eastern contour of the Antarctic Peninsula and the Antarctic continent up to 20◦E. The ecosystem of the Weddell Sea is one of the most pristine and relatively unexplored hotspots of marine biodiversity in Antarctica [@Teschke2021]. It contains several areas of ecological significance for birds and marine mammals, such as important breeding and foraging areas [@Hindell2020]. Furthermore, the sea floor is home to large and unique forest-like sponge and filter-feeder assemblages, which are particularly numerous on the continental shelf in the eastern part of the Weddell Sea. These species assemblages create a structurally complex and diverse habitat for thousands of species [@Barthel1992; @Brey1994; Brandt2007].

@@

<!-- Objective -->
The objective of this work was to improve the knowledge on how the Weddell Sea and the species therein may respond to perturbations from ongoing environmental changes. To achieve this we: 1) estimated the strength for each interaction in the Weddell Sea food web, 2) characterised species considering weighted and unweighted properties, and 3) analysed the species’ role in the stability of the food web.

@@

The Weddell Sea food web was retrieved from the GlobAL daTabasE of traits and food Web Architecture (GATEWAy, version 1.0) of the German Centre for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig [@Brose2018]. In addition to predator-prey interactions, the database contains information on other biological data such as the mean body mass and movement type for each species in the food web. Furthermore, it incorporates information about the interaction itself, such as the dimension of the predator search space (2 or 3 dimensions). In its current form the Weddell Sea food web comprises 490 species and 16041 predator-prey interactions and constitutes one of the most resolved food webs constructed to date [@Jacob2011].

@@

To estimate the strength of each pairwise interaction in the food web we followed the methodology proposed by @Pawar2012. The minimum data requirements are body mass of the consumer (predator) and resource (prey), and the interaction dimensionality (ID) classified as 2 or 3 dimensions. The ID is defined as the dimension of the search space of the predator, that is equivalent to  the movement space of the prey. Thus, the ID is classified as 2D when both predator and prey move in 2D (e.g., both are benthic) or if a predator moves in 3D and a prey in 2D (e.g., pelagic predator on benthic prey). The ID is classified as 3D when both predator and prey move in 3D (e.g., both pelagic) or if the predator moves in 2D and the prey in 3D (e.g., benthic predator, pelagic prey) [@Pawar2012]. GATEWAy v.1.0 provides information on the mean body mass for consumers and resources, except for ‘detritus’ and ‘sediment’, and the dimensionality for the majority of the interactions, though the latter is missing in some cases (924 interactions). To complete the missing data on species ‘dimensionality’, we used information about the movement type of predators and prey  included in GATEWAy.

@@

We obtained estimates for resource density and the search rate from the scaling relationships with the resource and the consumer mass, respectively [@Pawar2012]. The coefficients of such relationships, determined by ordinary least squares regression, vary with the interaction dimensionality. On one hand, resource density scales with resource mass as power-law with exponents  in 2D and  in 3D. Since mean mass for resources ‘phytodetritus’ and ‘sediment’ were not available in GATEWAy, we considered the body mass of the smallest phytoplankton species (‘Fragilariopsis cylindrus’) as a proxy. This is justified by the fact that ‘phytodetritus’ and ‘sediment’ are mainly composed of dead or senescent phytoplankton reaching the seabed [@Wolanski2011]. On the other hand, search rate scales with consumer mass as power-law with exponents in 2D and in 3D.

We fitted six candidate models (Exponential, Gamma, log-Normal, Normal, Power-law and Uniform) the interaction strength distribution using maximum likelihood [@McCallum2008], and selected the best fitting model by computing the Akaike Information Criterion AIC [@Burnham2002].

@@

To assess species roles as a function of the weighted food web, we focused on mean interaction strength defined as the average strength of all interactions for a given species. Further we calculated three unweighted species properties: a) species degree, i.e., the sum of in- and out-going interactions ; b) trophic level ; and c) trophic similarity, i.e., the trophic overlap based on shared and unique resources and consumers. These metrics were chosen to assess a species role based on the unweighted food web. The species degree has often been equated with species importance to the structure and functioning within a food web, i.e. perturbations to high-degree species may therefore have more significant effects on the food web robustness to perturbations than low-degree species [@Dunne2002a; references in @Cirtwill2018a]. The trophic level offers information about how important a species is to its biotic community, i.e., top predators and primary producers are expected to have particularly large effects on the rest of their communities through top-down and bottom-up control, respectively [references in @Cirtwill2018a]. Trophic similarity is an index of trophic overlap considering the set of prey and predators for a pair of species; it measures one of the most important aspects of species’ niches, the trophic niche, and functional aspects of biodiversity [@Martinez1991; @Williams2000].

Furthermore, we took a species’ habitat into account, which describes the physical position of a species within the ecosystem. Species were categorised as: 1) benthic, if a species lives on  the seafloor; 2) pelagic, if a species lives close to the surface; 3) benthopelagic, if it moves between and connects the mentioned environments; 4) demersal, if it lives and feeds on or near the bottom of the sea; and 5) land-based, if the consumer is not strictly aquatic but feeds predominantly on marine species. Species habitat affiliation was retrieved from @Jacob2011.

With the aim of studying the relationship between the interaction strength of the species (weighted property) and its unweighted properties we performed linear regression analyses between the log mean interaction strength and each of the mentioned unweighted properties. Thus we considered the interaction strength as the dependent variable and the given unweighted property as the independent variable, and obtained the coefficients (slope and intercept) for the linear model. Models were fitted using the least squares approach. We also explored the mean interaction strength distribution with the species habitat.

@@

To analyse the impact of species on food web stability, we performed extinction simulations deleting one species at a time, that is for every extinction network size was reduced by one species only. After each extinction, we calculated the stability of the network minus the targeted species (489 nodes) and compared it with that of the whole network (490 nodes in total). To calculate stability, we used the mean of the real part of the maximum eigenvalue of the Jacobian matrix using randomized Jacobians and keeping the predator-prey sign structure fixed [@Allesina2008; @Grilli2016]. This index indicates a more stable food web when it is negative. We performed 1000 simulations for each species extinction and obtained a mean maximum eigenvalue for each case. At last we statistically analysed such a difference with an Anderson-Darling test considering a p-value < 0.01 [@Scholz1987]. If this difference is positive, then the stability of the food web is higher if a species was removed, and vice versa. A detailed description on the stability calculations can be found in Supplementary Material.

Once we had the results for the impact on stability for each species extinction, we plotted them considering weighted (interaction strength) and unweighted properties, and species habitat. With this we aim to characterise those species with a relatively high effect on the stability of the food web.

@@

>>>>

<!-- Intro & advantages of studying food webs -->
Building a network of feeding interactions among species is a fundamental step to understanding ecological communities and is a valuable tool to manage ecosystems for conservation of biodiversity and functioning [@Thompson2012]. Although analyses considering presence/absence data on predator-prey interactions have provided insights about the structure and functioning of communities [e.g. @Kortsch2015; @Marina2018; @Cordone2020; @Rodriguez2022], more information is needed to effectively characterise the ecosystems' dynamics and stability. In this regard, the intensity of the interactions between species or the interaction strength became of paramount importance in recent years [@Carrara2015; @Allesina2015; @Nilsson2016].

<!-- Interaction strength -->
Several methodologies have been applyied to estimate the interaction strength in food webs, where the quantity and quality of data mostly determines which approach is more convenient [@Berlow2004]. <!-- Seems to me very short, it should follow a short description of the methodologies -->

<!-- General description of the Weddell Sea -->
The Weddell Sea represents the southerly part of the Atlantic Sector of the Southern Ocean. About one quarter of the Weddell Sea’s entire marine area lies over the continental shelf, stretching along the eastern contour of the Antarctic Peninsula and the Antarctic continent up to 20ºE. The ecosystem of the Weddell Sea is one of the most pristine and relatively unexplored hotspots of marine biodiversity in Antarctica [@Teschke2021]. It contains several areas of ecological significance for birds and marine mammals, such as important breeding and foraging areas [@Hindell2020]. Furthermore, the sea floor is home to large and unique forest-like sponge and filter-feeder assemblages, which are particularly numerous on the continental shelf in the eastern part of the Weddell Sea. These species assemblages create a structurally complex and diverse habitat for thousands of species [@Barthel1992; @Brey1994; Brandt2007].

@@

<!-- Objective -->
The objective of this work was to improve the knowledge of how the Weddell Sea and the species therein may respond to perturbations from ongoing environmental changes. To achieve this we: 1) estimated the strength of each interaction in the Weddell Sea food web, 2) characterised species considering weighted and unweighted properties, and 3) analysed the species’ role in the stability of the food web.

@@

The Weddell Sea food web was retrieved from the GlobAL daTabasE of traits and food Web Architecture (GATEWAy, version 1.0) of the German Centre for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig [@Brose2018]. In addition to predator-prey interactions, the database contains information on other biological data such as the mean body mass and movement type for each species in the food web. Furthermore, it incorporates information about the interaction itself, such as the dimension of the predator search space (2 or 3 dimensions). In its current form, the Weddell Sea food web comprises 490 species and 16041 predator-prey interactions and constitutes one of the most resolved food webs constructed to date [@Jacob2011].

@@

To estimate the strength of each pairwise interaction in the food web we followed the methodology proposed by @Pawar2012. The minimum data requirements are the body mass of the consumer (predator) and resource (prey), and the interaction dimensionality (ID) classified as 2 or 3 dimensions. The ID is defined as the dimension of the search space of the predator, that is equivalent to the movement space of the prey. Thus, the ID is classified as 2D when both predator and prey move in 2D (e.g., both are benthic) or if a predator moves in 3D and prey in 2D (e.g., a pelagic predator on benthic prey). The ID is classified as 3D when both predator and prey move in 3D (e.g., both pelagic) or if the predator moves in 2D and the prey in 3D (e.g., benthic predator, pelagic prey) [@Pawar2012]. GATEWAy v.1.0 provides information on the mean body mass for consumers and resources, except for ‘detritus’ and ‘sediment’, and the dimensionality for the majority of the interactions, though the latter is missing in some cases (924 interactions). To complete the missing data on species ‘dimensionality’, we used information about the movement type of predators and prey included in GATEWAy.

@@

We obtained estimates for resource density and the search rate from the scaling relationships with the resource and the consumer mass, respectively [@Pawar2012]. The coefficients of such relationships, determined by ordinary least squares regression, vary with the interaction dimensionality. On the one hand, resource density scales with resource mass as a power-law with different exponents in 2D and 3D. Since mean mass for resources ‘phytodetritus’ and ‘sediment’ were unavailable in GATEWAy, we considered the body mass of the smallest phytoplankton species (‘Fragilariopsis cylindrus’) as a proxy. This is justified by the fact that ‘phytodetritus’ and ‘sediment’ are mainly composed of dead or senescent phytoplankton reaching the seabed [@Wolanski2011]. On the other hand, the search rate scales with consumer mass as power-law with exponents in 2D and 3D.

After the estimation, we fitted six candidate models for the interaction strength distribution:  Exponential, Gamma, log-Normal, Normal, Power-law and Uniform. We used maximum likelihood [@McCallum2008] for the fitting, and selected the best model by computing the Akaike Information Criterion AIC [@Burnham2002].

@@

We calculated three unweighted species properties: a) species degree, i.e., the sum of in- and out-going interactions; b) trophic level; and c) trophic similarity, i.e., the trophic overlap based on shared and unique resources and consumers. These metrics were chosen to assess a species' role based on the unweighted food web. The species degree has often been equated with species' importance to the structure and functioning within a food web, i.e. perturbations to high-degree species may therefore have more significant effects on the food web robustness to perturbations than low-degree species [@Dunne2002a; references in @Cirtwill2018a]. The trophic level offers information about how important a species is to its biotic community, i.e., top predators and primary producers are expected to have particularly large effects on the rest of their communities through top-down and bottom-up control, respectively [references in @Cirtwill2018a]. The trophic similarity is an index of trophic overlap considering the set of prey and predators for a pair of species; it measures one of the most important aspects of species’ niches, the trophic niche, and functional aspects of biodiversity [@Martinez1991; @Williams2000].

Furthermore, we took a species’ habitat into account, which describes the physical position of a species within the ecosystem. Species were categorised as 1) benthic, if a species lives on the seafloor; 2) pelagic if a species lives close to the surface; 3) benthopelagic if it moves between and connects the mentioned environments; 4) demersal if it lives and feeds on or near the bottom of the sea; and 5) land-based if the consumer is not strictly aquatic but feeds predominantly on marine species. Species habitat affiliation was retrieved from @Jacob2011.

To assess species' roles as a function of the weighted food web, we focused on mean interaction strength defined as the average strength of all interactions for a given species. To study the relationship between the interaction strength of the species (weighted property) and its unweighted properties we performed linear regression analyses between the log mean interaction strength and each of the mentioned unweighted properties. Thus we considered the interaction strength as the dependent variable and the given unweighted property as the independent variable and obtained the coefficients (slope and intercept) for the linear model. Models were fitted using the least squares approach. We also explored the mean interaction strength distribution with the species' habitat.

@@

To analyse the impact of species on the food web stability, we performed extinction simulations deleting one species at a time, that is for every extinction network size was reduced by one species only. After each extinction, we calculated the stability of the network minus the targeted species (489 nodes) and compared it with that of the whole network (490 nodes in total). To calculate the stability, we used the mean of the real part of the maximum eigenvalue of the Jacobian matrix using randomized Jacobians and keeping the predator-prey sign structure fixed [@Allesina2008; @Grilli2016]. This index indicates a more stable food web when it is negative. We performed 1000 simulations for each species extinction and obtained a mean maximum eigenvalue for each case. At last, we statistically analysed such a difference with an Anderson-Darling test considering a p-value < 0.01 [@Scholz1987]. If this difference is positive, then the stability of the food web is higher if a species was removed, and vice versa. A detailed description of the stability calculations can be found in Supplementary Material.

Once we had the results for the impact on stability for each species extinction, we plotted them considering weighted (interaction strength) and unweighted properties, and species habitat. With this, we aim to characterise those species with a relatively high effect on the stability of the food web.

@@

<!--LAS up to here -->

### 2022-12-19 Manuscript/Manuscript.tex
LAS - Reviewed up to results

\introduction[Introduction]

@@

The statistical distribution that best fitted the empirical interaction
strength distribution was a `gamma' due to the high proportion of weak
interactions and the existence of a few strong interactions (Figure 3,
Table S3).

>>>>

\section{Introduction}

Building a network of feeding interactions among species is a
fundamental step to understand ecological communities and is a valuable
tool to manage ecosystems for conservation of biodiversity and
functioning \citep{Thompson2012}. Although analyses considering
presence/absence data on predator-prey interactions have provided
insights about the structure and functioning of communities
\citep[e.g.][]{Kortsch2015, Marina2018, Cordone2020, Rodriguez2022},
more information is needed to effectively charaterise the ecosystems'
dynamics and stability. In this regard, the intensity of the
interactions between species or the interaction strength became of
paramount importance in recent years
\citep{Carrara2015, Allesina2015, Nilsson2016}.

Several methodologies have been applyied to estimate the interaction
strength in food webs, where the quantity and quality of data mostly
determines which approach is more convenient \citep{Berlow2004}.

The Weddell Sea represents the southerly part of the Atlantic Sector of
the Southern Ocean. About one quarter of the Weddell Sea's entire marine
area lies over the continental shelf, stretching along the eastern
contour of the Antarctic Peninsula and the Antarctic continent up to
20◦E. The ecosystem of the Weddell Sea is one of the most pristine and
relatively unexplored hotspots of marine biodiversity in Antarctica
\citep{Teschke2021}. It contains several areas of ecological
significance for birds and marine mammals, such as important breeding
and foraging areas \citep{Hindell2020}. Furthermore, the sea floor is
home to large and unique forest-like sponge and filter-feeder
assemblages, which are particularly numerous on the continental shelf in
the eastern part of the Weddell Sea. These species assemblages create a
structurally complex and diverse habitat for thousands of species
{[}\citet{Barthel1992}; \citet{Brey1994}; Brandt2007{]}.

The Weddell Sea region is less affected by sea surface warming compared
with other areas of the Southern Ocean. If a climate change scenario is
considered, the Weddell Sea continental shelf will be probably a sink
area and a refuge for highly cold-adapted benthonic species
\citep{Griffiths2017}.

@@

The statistical distribution that best fitted the empirical interaction
strength distribution of the Weddell Sea food web was a `gamma' due to
the high proportion of weak interactions and the existence of a few
strong interactions (Figure 3, Table S3).
